Extended Formal Analysis of the EDHOC Protocol in Tamarin

### 2022-07-22 formalization.tex
[fix] heading consitency

\subsection{Formal Model}\label{sec:threat-model}
As in~\cite{Norr21}, we verify \mEdhoc{} in an extended
Dolev-Yao model~\cite{DY83}.
%
Dolev-Yao is a well-established model for symbolic verification of security
protocols.
%
Messages are modelled as terms in an algebra, and the various cryptographic
operations are assumed to be perfect, e.g., encrypted messages can only be
decrypted with the correct key, and there are no hash collisions.
%

@@

%----------------------------------------------------------------------- PFS
\subsubsection{Secrecy of the Session-key Material.}
\label{sec:secrecy}
We show that the adversary cannot gain access to the session-key material
\mSessKey{}, even under a weak post-compromise security model that allows the
adversary limited access to the long-term keys of the parties involved through a
secure interface before the session starts, unrestricted access to the long-term
keys of the involved parties after the session ends and of all other parties, as
well as access to all other session keys.
%
This property is formalized as \mPredPcs{} in Figure~\ref{fig:props}.
%

@@

%------------------------------------------------------- Agreed parameters
\subsubsection{Parameters in set $S$.}
\label{sec:agreedParams}
We now describe the set $S$ of parameters upon which the two parties obtain
guarantees via the above properties.
%
The initiator $I$ gets injective and implicit agreement guarantees on the
following partial set $S_P$ of parameters~\cite{Norr21}:
\begin{itemize}
    \item the roles played by itself and its peer,
    \item responder identity,
    \item session-key material (which varies depending on the \mEdhoc{} 
method),
    \item context identifiers \mCi{} and \mCr{}, and
    \item cipher suites \mSuites{}.
\end{itemize}
%

@@

%---------------------------------------------------------------------------sub
\subsection{Encoding \mEdhoc{} in \mTamarin}
\label{sec:modeling}
%
In this section, we describe how we model the \mSigSig, \mSigStat, \mStatSig,
and \mStatStat{} methods of \mEdhoc{} in \mTamarin.
%
As in~\cite{Norr21}, we construct the \mTamarin{} model by utilizing the fact
that all methods of \mEdhoc{} share a common underlying structure
(as shown in Figure~\ref{fig:edhocFramework}).
%
We do so by using a single specification file in the M4 macro language,
which generates all the methods.
%
As in~\cite{Norr21}, we only present the \mStatSig{} method which illustrates
the use of two different authentication methods.
%
The full \mTamarin{} code for all models can be found 
at~\cite{edhocTamarinRepo}.
%

@@

Having described how we adapt the equational theory to model \mEdhoc,
we now move on to the modelling of the adversary model and the 
environment in
which the protocol is executed.
%
We extend the built-in Dolev-Yao adversary rules which are part of 
\mTamarin.
%
We use the following rules to capture the link between a party's identity and
their long-term key pairs, in the \mSig{}- and in the \mStat{}-based methods
respectively.
\begin{center}
\begin{minipage}{0.48\textwidth}
\begin{scriptsize}
\begin{verbatim}
1 rule registerLTK_SIG:
2    [Fr(~ltk)] --[UniqLTK($A, ~ltk)]->
3        [!LTK_SIG($A, ~ltk),
4         !PK_SIG($A, pk(~ltk)),
5         Out(<$A, pk(~ltk)>)]
\end{verbatim}
\end{scriptsize}
\end{minipage}
\hfill\vline\hfill
\begin{minipage}{0.48\textwidth}
\begin{scriptsize}
\begin{verbatim}
1 rule registerLTK_STAT:
2    [Fr(~ltk)] --[UniqLTK($A, ~ltk)]->
3        [!LTK_STAT($A, ~ltk),
4         !PK_STAT($A, 'g'^~ltk),
5         Out(<$A, 'g'^~ltk>)]
\end{verbatim}
\end{scriptsize}
\end{minipage}
\end{center}

>>>>

%---------------------------------------------------------------------- subsub
\subsection{Formal Model}\label{sec:threat-model}
As in~\cite{Norr21}, we verify \mEdhoc{} in an extended
Dolev-Yao model~\cite{DY83}.
%
Dolev-Yao is a well-established model for symbolic verification of security
protocols.
%
Messages are modelled as terms in an algebra, and the various cryptographic
operations are assumed to be perfect, e.g., encrypted messages can only be
decrypted with the correct key, and there are no hash collisions.
%

@@

%----------------------------------------------------------------------- PFS
\subsubsection{Secrecy of the Session-Key Material.}
\label{sec:secrecy}
We show that the adversary cannot gain access to the session-key material
\mSessKey{}, even under a weak post-compromise security model that allows the
adversary limited access to the long-term keys of the parties involved through a
secure interface before the session starts, unrestricted access to the long-term
keys of the involved parties after the session ends and of all other parties, as
well as access to all other session keys.
%
This property is formalized as \mPredPcs{} in Figure~\ref{fig:props}.
%

@@

%------------------------------------------------------- Agreed parameters
\subsubsection{Set $S$ of Agreed Parameters.}
\label{sec:agreedParams}
We now describe the set $S$ of parameters upon which the two parties obtain
guarantees via the above properties.
%
The initiator $I$ gets injective and implicit agreement guarantees on the
following partial set $S_P$ of parameters~\cite{Norr21}:
\begin{itemize}
    \item the roles played by itself and its peer,
    \item responder identity,
    \item session-key material (which varies depending on the \mEdhoc{} 
method),
    \item context identifiers \mCi{} and \mCr{}, and
    \item cipher suites \mSuites{}.
\end{itemize}
%

@@

%---------------------------------------------------------------------------sub
\subsection{Encoding \mEdhoc{} in \mTamarin{}}
\label{sec:modeling}
%
In this section, we describe how we model the \mSigSig, \mSigStat, \mStatSig,
and \mStatStat{} methods of \mEdhoc{} in \mTamarin.
%
As in~\cite{Norr21}, we construct the \mTamarin{} model by utilizing the fact
that all methods of \mEdhoc{} share a common underlying structure
(as shown in Figure~\ref{fig:edhocFramework}).
%
We do so by using a single specification file in the M4 macro language,
which generates all the methods.
%
As in~\cite{Norr21}, we only present the \mStatSig{} method which illustrates
the use of two different authentication methods.
%
The full \mTamarin{} code for all models can be found 
at~\cite{edhocTamarinRepo}.
%

@@

Having described how we adapt the equational theory to model \mEdhoc,
we now move on to the modeling of the adversary model and the 
environment in
which the protocol is executed.
%
We extend the built-in Dolev-Yao adversary rules which are part of 
\mTamarin.
%
We use the following rules to capture the link between a party's identity and
their long-term key pairs, in the \mSig{}- and in the \mStat{}-based methods
respectively.
\begin{center}
\begin{minipage}{0.48\textwidth}
\begin{scriptsize}
\begin{verbatim}
1 rule registerLTK_SIG:
2    [Fr(~ltk)] --[UniqLTK($A, ~ltk)]->
3        [!LTK_SIG($A, ~ltk),
4         !PK_SIG($A, pk(~ltk)),
5         Out(<$A, pk(~ltk)>)]
\end{verbatim}
\end{scriptsize}
\end{minipage}
\hfill\vline\hfill
\begin{minipage}{0.48\textwidth}
\begin{scriptsize}
\begin{verbatim}
1 rule registerLTK_STAT:
2    [Fr(~ltk)] --[UniqLTK($A, ~ltk)]->
3        [!LTK_STAT($A, ~ltk),
4         !PK_STAT($A, 'g'^~ltk),
5         Out(<$A, 'g'^~ltk>)]
\end{verbatim}
\end{scriptsize}
\end{minipage}
\end{center}

### 2022-07-22 paper.tex
[fix] heading consitency

\begin{abstract}
Given how common IoT devices that use constrained resources are becoming
today, the need of the hour is communication protocols which can operate
securely under such limitations.
%
For a few years, the Internet Engineering Task Force (IETF) has been working
to standardize \mEdhoc{}, an authenticated key establishment protocol for
such constrained IoT devices.
%
The first version of \mEdhoc{} was proposed in 2016.
%
In 2018, Bruni et al.~\cite{DBLP:conf/secsr/BruniJPS18} used the \mProverif{}
tool~\cite{DBLP:conf/csfw/Blanchet01} to formally analyze an early version of
\mEdhoc{}, which had only two key establishment methods.
%
%By now, the protocol has been fleshed out much more, and has been
%augmented with multiple new key establishment methods.
By 2021, the protocol had been fleshed out much more, with multiple new key
establishment methods, and this version was formally analyzed using the
\mTamarin{} prover~\cite{DBLP:conf/cav/MeierSCB13} in~\cite{Norr21}.
%
Here, we build on this work, by modifying the model, adding new properties
that we analyze, and discussing some of the aspects of the latest \mEdhoc{}
specification. In particular, we extend the modelling with trusted execution
environments (TEEs), modify the way we model \mXor{} encryption, and
verify weak post-compromise security (PCS) as well as the secrecy and
integrity of some additional data used as part of the protocol.
\keywords{Formal Verification, Symbolic Dolev-Yao Model,
           Authenticated Key Establishment, Protocols, IoT, Tamarin.}

@@

%-------------------------------------------------------------------------- sec
\section{\uppercase{Error handling}}
\label{sec:errorHandling}
Section 6 of the current \mEdhoc{}
specification~\cite{draft-ietf-lake-edhoc-15}
states that error messages can be sent at any time and by any party.
%
There are three types of error messages: type 0 is used to represent success,
type 1 represents a generic error message used for debugging purposes, and
type 2 represents the failure to negotiate a common ciphersuite.
%
Type 2 messages carry a list of supported ciphersuites by the responder.
%
The contents and semantics of type 1 messages are dependent on the
implementation.
%
Consequently, type 1 error messages provide a generic message passing
mechanism, albeit without a predefined semantics.
%
We now discuss some issues with this specification of error messages.
%
In this section, we use ``specification'' to mean the latest specification, as
in~\cite{draft-ietf-lake-edhoc-15}.

\subsection{Lack of connection identifiers}
Error messages do not include the optional connection identifiers \mCi{} or
\mCr{}.
%
This is potentially problematic for connection-less transport layers, where
these identifiers may be used to correlate messages.

\subsection{Improper handling}
\mEdhoc{} provides an error message mechanism, but little to no guidance on
how it should be used safely.
%
For example, assume that an application using \mEdhoc{} logs error messages
on a finite log with rotation.
%
If the log is used for anomaly detection or detection of sensitive events, then
an adversary can simply inject error messages and fill the log until the
sensitive event is overwritten.
%
In such situations, implementors should do proper log-separation and log
management.

\subsection{Unspecified semantics for debugging messages}
Section 6.2 gives an example of sending a type 1 error message with the
information string ``Method not supported".
%
Note that \mEdhoc{} does not include a mechanism to support the
negotiating of the connection method.
%
The initiator selects the method as part of the first message, and if the
responder replies with a type 1 error message along with the information
string ``Method not supported", the standard does not tell an
implementation how to act upon this information string for negotiation
purposes.
%
An adversary can fake an error message, and there is no algorithm that is
specified to ensure that a common available authentication mechanism is
always selected.
%
In particular, note that Section 5.1 in the specification states that if processing
fails for some reason, then ``typically" an error message is sent, processing is
terminated, and protocol state is discarded.
%
But negotiation needs us to remember state.

With respect to ciphersuite negotiation, Section 5.2.2 of the specification also
states ``The Initiator MUST NOT change the supported cipher suites and
the order of preference in SUITES\_I based on previous error messages.''
%
However, \mSuites is sent in the clear, and therefore can be modified by the
adversary.
%
It is important to consider what happens if the order of preference in
\mSuites{} is changed either by the adversary, or by the initiator themselves
for some other reason.
%
It might be helpful to include a flag in \mMsgone{} to indicate to the
responder
that a (modified) ciphersuite ordering depends on that from a previous
execution.

%-------------------------------------------------------------------------- sec
\section{\uppercase{Improvements to the specification}}
\label{sec:discussion}
The \mEdhoc{} \mSpec{} has many points where it can be improved.
%
We discovered a few of these issues in~\cite{Norr21}, but in this work,
we also discuss a couple of new issues.
%
We communicated the points discussed in~\cite{Norr21} to the authors of the
\mSpec{}.
%
Here, we also list the protocol developers' response (if any) to being told of these
ambiguities in the \mSpec{}.

>>>>

\begin{abstract}
Given how common IoT devices that use constrained resources are becoming
today, the need of the hour is communication protocols which can operate
securely under such limitations.
%
For a few years, the Internet Engineering Task Force (IETF) has been working
to standardize \mEdhoc{}, an authenticated key establishment protocol for
such constrained IoT devices.
%
The first version of \mEdhoc{} was proposed in 2016.
%
In 2018, Bruni et al.~\cite{DBLP:conf/secsr/BruniJPS18} used the \mProverif{}
tool~\cite{DBLP:conf/csfw/Blanchet01} to formally analyze an early version of
\mEdhoc{}, which had only two key establishment methods.
%
%By now, the protocol has been fleshed out much more, and has been
%augmented with multiple new key establishment methods.
By 2021, the protocol had been fleshed out much more, with multiple new key
establishment methods, and this version was formally analyzed using the
\mTamarin{} prover~\cite{DBLP:conf/cav/MeierSCB13} in~\cite{Norr21}.
%
Here, we build on this work, by modifying the model, adding new properties
that we analyze, and discussing some of the aspects of the latest \mEdhoc{}
specification. In particular, we extend the modeling with trusted execution
environments (TEEs), modify the way we model \mXor{} encryption, and
verify weak post-compromise security (PCS) as well as the secrecy and
integrity of some additional data used as part of the protocol.
\keywords{Formal Verification, Symbolic Dolev-Yao Model,
           Authenticated Key Establishment, Protocols, IoT, Tamarin.}

@@

%-------------------------------------------------------------------------- sec
\section{\uppercase{Error Handling}}
\label{sec:errorHandling}
Section 6 of the current \mEdhoc{}
specification~\cite{draft-ietf-lake-edhoc-15}
states that error messages can be sent at any time and by any party.
%
There are three types of error messages: type 0 is used to represent success,
type 1 represents a generic error message used for debugging purposes, and
type 2 represents the failure to negotiate a common ciphersuite.
%
Type 2 messages carry a list of supported ciphersuites by the responder.
%
The contents and semantics of type 1 messages are dependent on the
implementation.
%
Consequently, type 1 error messages provide a generic message passing
mechanism, albeit without a predefined semantics.
%
We now discuss some issues with this specification of error messages.
%
In this section, we use ``specification'' to mean the latest specification, as
in~\cite{draft-ietf-lake-edhoc-15}.

\subsection{Lack of Connection Identifiers}
Error messages do not include the optional connection identifiers \mCi{} or
\mCr{}.
%
This is potentially problematic for connection-less transport layers, where
these identifiers may be used to correlate messages.

\subsection{Proper Reception Handling}
\mEdhoc{} provides an error message mechanism, but little to no guidance on
how it should be used safely.
%
For example, assume that an application using \mEdhoc{} logs error messages
on a finite log with rotation.
%
If the log is used for anomaly detection or detection of sensitive events, then
an adversary can simply inject error messages and fill the log until the
sensitive event is overwritten.
%
In such situations, implementers should do proper log-separation and log
management.

\subsection{Unspecified Semantics for Debugging Messages}
Section 6.2 gives an example of sending a type 1 error message with the
information string ``Method not supported".
%
Note that \mEdhoc{} does not include a mechanism to support the
negotiating of the connection method.
%
The initiator selects the method as part of the first message, and if the
responder replies with a type 1 error message along with the information
string ``Method not supported", the standard does not tell an
implementation how to act upon this information string for negotiation
purposes.
%
An adversary can fake an error message, and there is no algorithm that is
specified to ensure that a common available authentication mechanism is
always selected.
%
In particular, note that Section 5.1 in the specification states that if processing
fails for some reason, then ``typically" an error message is sent, processing is
terminated, and protocol state is discarded.
%
But negotiation needs us to remember state.

With respect to ciphersuite negotiation, Section 5.2.2 of the specification also
states ``The Initiator MUST NOT change the supported cipher suites and
the order of preference in SUITES\_I based on previous error messages.''
%
However, \mSuites{} is sent in the clear, and therefore can be modified by the
adversary.
%
It is important to consider what happens if the order of preference in
\mSuites{} is changed either by the adversary, or by the initiator themselves
for some other reason.
%
It might be helpful to include a flag in \mMsgone{} to indicate to the
responder
that a (modified) ciphersuite ordering depends on that from a previous
execution.

%-------------------------------------------------------------------------- sec
\section{\uppercase{Improvements to the Specification}}
\label{sec:discussion}
The \mEdhoc{} \mSpec{} has many points where it can be improved.
%
We discovered a few of these issues in~\cite{Norr21}, but in this work,
we also discuss a couple of new issues.
%
We communicated the points discussed in~\cite{Norr21} to the authors of the
\mSpec{}.
%
Here, we also list the protocol developers' response (if any) to being told of these
ambiguities in the \mSpec{}.
